Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-08-22 manuscript.qmd
pushing revised manuscript

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- or electro-encephalography (M/EEG) offer a unique window onto the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time and space, followed by corrections for multiple comparisons such as cluster-based inference. While effective for controlling error rates at the cluster-level, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual time points to clusters, it prevents drawing conclusions about the precise onset or offset of observed effects. Here, we introduce a *model-based* approach for analysing M/EEG timeseries such as event-related potentials (ERPs) or decoding performance over time. Our approach leverages Bayesian generalised additive multilevel models, providing posterior probabilities that an effect is above zero (or above chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and actual M/EEG datasets, we demonstrate that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable results. We provide an R package implementing the method and describe how it can be integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, cluster-based inference, simulation, multiple comparisons, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# see available font via systemfonts::system_fonts()
# monofont: Monaco
# monofont: Courier
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
suppress-title-introduction: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
          \usepackage{booktabs, caption, longtable, colortbl, array}
          \usepackage{setspace}
          \onehalfspacing
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(tidytext)
library(grateful)
library(ggrepel)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)
library(mgcv)
library(gt)

# setting knitr options
knitr::opts_chunk$set(
    cache = TRUE,
    eval = TRUE,
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    out.width = "100%",
    fig.asp = 0.75,
    fig.pos = "!htb",
    dev = "png",
    dpi = 300
    )

# disabling hyperlinks
# options(cli.hyperlink = FALSE)

@@

@@

The conventional approach involves mass-univariate statistical testing through time and/or space followed by some form of correction for multiple comparisons with the goal of maintaining the familywise error rate (FWER) or the false discovery rate (FDR) at the nominal level (e.g., 5%). Cluster-based inference is the most common way of achieving this sort of error control in the M/EEG literature, being the recommended approach in several software programs [e.g., `EEGlab`, @delorme2004; `MNE-Python`, @gramfort2013]. While effective for controlling error rates, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it prevents the ability to draw precise conclusions about the spatiotemporal localisation of such effects [@maris2007; @sassenhagen2019]. As pointed out by @maris2007: "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA [false alarm] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect". Even worse, @rosenblatt2018 note that cluster-based inference suffers from low spatial resolution: "Since discovering a cluster means that 'there exists at least one voxel with an evoked response in the cluster', and not that 'all the voxels in the cluster have an evoked response', it follows that the larger the detected cluster, the less information we have on the location of the activation." As a consequence, cluster-based inference is expected to perform poorly for localising the onset of M/EEG effects; a property that was later demonstrated in simulation studies [e.g., @rousselet_using_2025; @sassenhagen2019].

To overcome the limitations of cluster-based inference, we introduce a novel *model-based* approach for precisely localising M/EEG effects in time, space, and other dimensions. The proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the posterior probability of effects being above chance at the level of timesteps, sensors, voxels, etc, while naturally taking into account spatiotemporal dependencies present in M/EEG data. We compare the performance of the proposed approach to well-established alternative methods using both simulated and actual M/EEG data and show that it significantly outperforms alternative methods in estimating the onset and offset of M/EEG effects.

@@

The issues with multiple comparisons represent a common and well-recognised danger in neuroimaging and M/EEG research, where the collected data allows for a multitude of potential hypothesis tests and is characterised by complex structures of spatiotemporal dependencies. The probability of obtaining at least one false positive in an ensemble (family) of $m$ tests (i.e., the FWER) is computed as $1-\left(1-\alpha\right)^{m}$ (for $m=10$ tests and $\alpha=0.05$, it is approximately equal to $0.4$). Different methods exist to control the FWER (i.e., bring it back to $\alpha$). Most methods apply a simple correction to series of $p$-values issued from univariate statistical tests (e.g., t-tests). For instance, the Bonferroni correction [@dunn1961] consists in setting the significance threshold to $\alpha/m$, or equivalently, multiplying the $p$-values by $m$ and using the standard $\alpha$ significance threshold. This method is generally overconservative (i.e., under-powered) as it assumes statistical independence of the tests, an assumption that is clearly violated in the context of M/EEG timeseries characterised by massive spatiotemporal dependencies. Some alternative methods aims at controlling the FDR, defined as the proportion of false positive *among positive tests* [e.g., @benjamini1995; @benjamini2001]. However, a major limitation of both types of corrections is that they do not take into account the spatial and temporal information contained in M/EEG data.

A popular technique to account for spatiotemporal dependencies while controlling the FWER is cluster-based inference [@bullmore1999; @maris2007]. A typical cluster-based inference consists of two successive steps [for more details on cluster-based inference, see for instance @frossard2022; @maris2011; @maris2007; @sassenhagen2019]. First, clusters are defined as sets of contiguous timesteps, sensors, voxels, etc, whose activity, summarised by some test statistic (e.g., a $t$-value), exceeds a predefined threshold (e.g., the 95th percentile of the parametric null distribution). Clusters are then characterised by their height (i.e., maximal value), extent (number of constituent elements), or some combination of both, for instance by summing the statistics within a cluster, an approach referred to as "cluster mass" [@maris2007; @pernet2015]. Then, the null hypothesis is tested by computing a $p$-value for each identified cluster by comparing its mass with the null distribution of cluster masses (obtained via permutation). As alluded previously, a significant cluster is a cluster which contains *at least one* significant time-point. As such, it would be incorrect to conclude, for instance, that the timestep of a significant cluster is the first moment at which some conditions differ [@frossard2022; @sassenhagen2019]. In other words, because the inference is performed at the second step (i.e., once clusters have been formed), no conclusion can be made about individual datapoints (e.g., timesteps, sensors, etc).

@@

where $h_{0}$ is typically $0$ and parameters $E$ and $H$ are set a priori (typically to 0.5 and 2, respectively) and control the influence of the extend and height on the TFCE. Note that in practice, this integral is approximated by a sum over small $h$ increments. Then, a *p*-value for each timestep $t$ is computed by comparing its TFCE with the null distribution of TFCE values (obtained via permutation). For each permuted signal, we keep the maximal value over the whole signal for the null distribution of the TFCE. The TFCE combined with permutation (assuming a large enough number of permutations) has been shown to provide accurate FWER control [e.g., @pernet2015]. However, further simulation work showed that cluster-based methods (including TFCE) perform poorly in localising the onset of M/EEG effects [e.g., @rousselet_using_2025; @sassenhagen2019].

@@

Scalp-recorded M/EEG signals capture neural activity originating from various brain regions and are often contaminated by artifacts unrelated to the cognitive processes under investigation. Consequently, analysing M/EEG data necessitates methods that can disentangle task-relevant neural signals from extraneous "noise." A widely adopted technique for this purpose is the estimation of event-related potentials (ERPs), which are stereotyped electrophysiological responses time-locked to specific sensory, cognitive, or motor events. Typically, ERPs are derived by averaging EEG or MEG epochs across multiple trials aligned to the event of interest (e.g., stimulus onset), thereby enhancing the signal-to-noise ratio by attenuating non-time-locked activity. However, this averaging approach has notable limitations: it assumes consistent latency and amplitude across trials and is primarily suited for simple categorical designs. Such assumptions may not hold in more complex experimental paradigms, potentially leading to suboptimal ERP estimations.

To overcome the limitations of averaging, a number of model-based options for estimating ERPs have been proposed. These approaches are justified by the observation that the averaging approach to ERPs is equivalent to an intercept-only linear regression in a simple categorical design with no overlapping factors [@smith2014a]. The regression approach to ERPs is much more flexible than simpl averaging, as it allows modelling linear and non-linear effects of continuous predictors such as word frequency or age [e.g., @smith2014a; @smith2014b; @tremblay2014] and allow modelling and disentangling overlapping processes [e.g., @skukies_brain_2024; @skukies_modelling_2021] as implemented in the UNFOLD toolbox [@ehinger_unfold_2019]. The regression approach, as implemented for instance in the LIMO toolbox [@pernet2011], is a multi-stage procedure in which a participant-level regression model is fitted independently to each datapoint (e.g., channel, timestep, etc.) to estimate the ERP, followed by group-level statistical analyses of the regression coefficients, itself followed by multiplicity corrections or cluster-based inference [for recent applied examples, see for instance @dunagan2025; @wüllhorst2025]. Whereas this approach allows including linear and non-linear discrete and continuous predictors when estimating the ERP, it has several limitations. First, the participant-level modelling ignores spatiotemporal dependencies as it treats all datapoints independently. Second, this approach does not take into account the participant (or stimulus) variability when testing group-level hypotheses (as could be done for instance via multilevel modelling). Finally, because the output of this procedure is summarised by cluster-based inference, its conclusions suffer from the same limitations discussed in the previous section.

To overcome the limitations of simple averaging, several model-based approaches for estimating ERPs have been proposed. These methods are motivated by the observation that traditional ERP averaging is mathematically equivalent to fitting an intercept-only linear regression model in a simple categorical design without overlapping events [@smith2014a]. In contrast to simple averaging, regression-based approaches to ERP estimation offer substantially greater flexibility. Notably, they allow for the modelling of both linear and nonlinear effects of continuous predictors--such as word frequency or age [e.g., @smith2014a; @smith2014b; @tremblay2014]--and enable the disentangling of overlapping cognitive processes [e.g., @ehinger_unfold_2019; @skukies_brain_2024; @skukies_modelling_2021]. One widely used implementation of this approach is provided by the LIMO EEG toolbox [@pernet2011], which follows a multi-stage analysis pipeline. First, a separate regression model is fit for each datapoint (e.g., each time point and electrode) at the individual participant level to estimate ERP responses. This is followed by group-level statistical analyses of the resulting regression coefficients, often accompanied by multiple comparison corrections or cluster-based inference [for recent applied examples, see @dunagan2025; @wüllhorst2025].

Although this framework allows for the inclusion of a wide range of predictors--both continuous and categorical, linear and nonlinear--it still has important limitations. First, fitting separate models for each datapoint ignores the spatiotemporal dependencies inherent in M/EEG data, potentially reducing statistical power and interpretability. Second, the subsequent group-level analyses typically do not account for hierarchical dependencies, between-participant or between-stimulus variability, which could otherwise be addressed through multilevel modelling. Finally, because the output of this procedure is summarised by cluster-based inference, its conclusions remain subject to the limitations discussed in the previous section.

Building upon the regression framework, recent advances allow modelling and disentangling overlapping processes [e.g., @skukies_brain_2024; @skukies_modelling_2021] as implemented in the UNFOLD toolbox [@ehinger_unfold_2019]...

In addition to modelling the non-linear effects of continuous predictors (e.g., word frequency) on the ERP amplitude, GAMs have been used for modelling the effect of time, that is, the shape of the ERP itself [@abugaber2023; @baayen2018; @meulman2015; @meulman2023]... In the next section, we turn to a brief introduction to generalised additive models and discuss their use for modelling M/EEG timeseries...

Beyond modelling nonlinear effects of continuous predictors on ERP amplitudes, generalised additive models (GAMs) have been employed to capture the temporal dynamics of ERPs themselves, effectively modelling the shape of the waveform over time [@abugaber2023; @baayen2018; @meulman2015; @meulman2023]. This approach allows for the estimation of smooth, data-driven functions that characterise how neural responses evolve over time, offering a flexible alternative to traditional linear models. In the following section, we provide a brief introduction to GAMs, highlighting their applicability to M/EEG time series analysis and the advantages they offer over conventional methods.

@@

In generalised additive models (GAMs), the functional relationship between the predictors and the response variable is decomposed into a sum of low-dimensional non-parametric functions. A typical GAM has the following form:

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this point: *splines* are functions composed of simpler functions. These simpler functions are called *basis functions* (e.g., cubic polynomial, thin-plate) and the set of basis functions is called a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions (@fig-intro-gam\ignorespaces A). Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function [@miller2025; @wood2017]. In other words, more complex (wiggly) basis functions are automatically penalised.

@@

```{r fig-intro-gam, echo = FALSE, fig.width = 9, fig.cap = "Different types of GAMs. **A**: GAMs predictions are computed as the weigthed sum (in black) of basis functions (here thin-plate basis functions, in grey). **B**: Constant-effect GAM, with 5 participants in colours and the group-level prediction in black. **C**: Varying-intercept + varying-slope GAMM (with constant smoother). **D**: Varying-intercept + varying-slope + varying-smoother GAMM. In this model, each participant gets its own intercept, slope, and degree of 'wiggliness' (smoother)."}
########################
# simulating some data #
########################

@@

A detailed treatment of the technical underpinnings of GAMs is beyond the scope of this article [see reference books such as @hastie2017; @wood2017]. However, it is worth emphasising that GAMs have been successfully applied to a wide range of time series data across the cognitive sciences, including pupillometry [e.g., @vanrij2019], articulography [e.g., @wieling2018], speech formant dynamics [e.g., @sóskuthy2021], neuroimaging data [e.g., @dinga2021], and event-related potentials [e.g., @abugaber2023; @baayen2018; @meulman2015; @meulman2023]. Their appeal for modelling M/EEG data lies in their ability to flexibly capture the complex shape of ERP waveforms without overfitting, through the use of smooth functions constrained by penalisation. Recent extensions, such as distributional GAMs [@rigby2005; @umlauf2018], allow researchers to model not only the mean structure but also the variance (or scale) and other distributional properties as functions of predictors--a feature that has proven useful in modelling neuroimaging data [e.g., @dinga2021]. Moreover, hierarchical or multilevel GAMs [@pedersen_hierarchical_2019] provide a principled way to account for the nested structure of M/EEG data (e.g., trials within participants), enabling the inclusion of varying intercepts, slopes, and smoothers (as illustrated in @fig-intro-gam\ignorespaces C-D). This approach mitigates the risk of overfitting and reduces the influence of outliers on smooth estimates [@baayen2020; @meulman2023].

In the following, we introduce our proposed approach based on Bayesian generalized additive multilevel models (Bayesian GAMMs), which builds on these strengths while providing fully probabilistic estimates of ERP dynamics. We illustrate how this framework enables direct inference on the presence and timing of neural effects at the trial level, across conditions and participants, and show that it outperforms conventional methods in both simulated and real M/EEG datasets.

@@

Cluster-based permutation tests are widely used in M/EEG research to identify statistically significant effects across time and space. However, these methods have notable limitations, particularly in accurately determining the precise onset and offset of neural effects. To address these limitations, we developed a model-based approach relying on Bayesian generalised additive multilevel models implemented in `R` via the `brms` package [@brms2017; @brms2018]. We evaluated the performance of this approach against conventional methods using both simulated and actual M/EEG data. Our findings demonstrate that this method provides more precise and reliable estimates of effect onsets and offsets, offering a valuable tool for researchers seeking to enhance the temporal resolution of their analyses.

# Methods

## M/EEG data simulation

Following the approach of @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see code in the online supplementary materials and details in @yeung2004]. As in @rousselet_using_2025, the signal was generated from a truncated Gaussian distribution with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated data for a group of 20 participants (with variable true onset) with 50 trials per participant and condition (@fig-eeg). All figures and simulation results can be reproduced using the `R` code available on GitHub: <https://github.com/lnalborczyk/brms_meeg>.

@@

## Model description and model fitting

>>>>

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- and electroencephalography (M/EEG) offer a unique window onto the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time and space followed by corrections for multiple comparisons or some form of cluster-based inference. While effective for controlling error rates at the cluster-level, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual time points to clusters, it prevents drawing conclusions about the precise onset or offset of observed effects. Here, we present a *model-based* alternative for analysing M/EEG timeseries, such as event-related potentials or time-resolved decoding accuracy. Our approach leverages Bayesian generalised additive multilevel models, providing posterior odds that an effect exceeds zero (or chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and empirical M/EEG datasets, we show that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable estimates. We provide an open-source R package implementing the method and describe how it can be integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, cluster-based inference, multiple comparisons, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
word-count: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# see available font via systemfonts::system_fonts()
# monofont: Monaco
# monofont: Courier
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
number-sections: true
number-depth: 3
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    # number_sections: true
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
          \usepackage{booktabs, caption, longtable, colortbl, array}
          \usepackage{setspace}
          \onehalfspacing
---

```{r setup, include = FALSE}
# loading packages
library(changepoint)
library(tidyverse)
library(tidybayes)
library(patchwork)
library(MetBrewer)
library(tidytext)
library(grateful)
library(ggrepel)
library(ggpubr)
library(scales)
library(pakret)
library(scico)
library(knitr)
library(brms)
library(mgcv)
library(gt)

# setting knitr options
knitr::opts_chunk$set(
    cache = TRUE,
    eval = TRUE,
    echo = FALSE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    out.width = "100%",
    fig.asp = 0.75,
    fig.pos = "!htb",
    dev = "png",
    dpi = 300
    )

@@

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}

@@

The conventional approach involves mass-univariate statistical testing through time and/or space followed by some form of correction for multiple comparisons with the goal of maintaining the familywise error rate (FWER) or the false discovery rate (FDR) at the nominal level (e.g., 5%). Cluster-based inference is the most common way of achieving this sort of error control in the M/EEG literature, being the recommended approach in several software programs [e.g., `EEGlab`, @delorme2004; `MNE-Python`, @gramfort2013]. While effective for controlling error rates, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it prevents the ability to draw precise conclusions about the spatiotemporal localisation of such effects [@maris2007; @sassenhagen2019]. As pointed out by @maris2007: "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA [false alarm] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect". Even worse, @rosenblatt2018 note that cluster-based inference suffers from low spatial resolution: "Since discovering a cluster means that 'there exists at least one voxel with an evoked response in the cluster', and not that 'all the voxels in the cluster have an evoked response', it follows that the larger the detected cluster, the less information we have on the location of the activation." As a consequence, cluster-based inference is expected to perform poorly for identifying the onset of M/EEG effects; a property that was later demonstrated in simulation studies [e.g., @rousselet_using_2025; @sassenhagen2019].

To overcome the limitations of cluster-based inference, we introduce a novel *model-based* approach for precisely localising M/EEG effects in time, space, and other dimensions. The proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the posterior odds of effects being above chance at the level of timesteps, sensors, voxels, etc, while naturally taking into account spatiotemporal dependencies present in M/EEG data. We compare the performance of the proposed approach to well-established alternative methods using both simulated and actual M/EEG data and show that it significantly outperforms alternative methods in estimating the onset and offset of M/EEG effects.

@@

The issues with multiple comparisons represent a common and well-recognised danger in neuroimaging and M/EEG research, where the collected data allows for a multitude of potential hypothesis tests and is characterised by complex structures of spatiotemporal dependencies. The probability of obtaining at least one false positive in an ensemble (family) of $m$ tests (i.e., the FWER) is computed as $1-\left(1-\alpha\right)^{m}$ (for $m=10$ independent tests and $\alpha=0.05$, it is approximately equal to $0.4$). Different methods exist to control the FWER, that is, to bring it back to $\alpha$. Most methods apply a simple correction to series of $p$-values issued from univariate statistical tests (e.g., t-tests). For instance, the Bonferroni correction [@dunn1961] consists in setting the significance threshold to $\alpha/m$, or equivalently, multiplying the $p$-values by $m$ and using the standard $\alpha$ significance threshold. This method is generally overconservative (i.e., under-powered) as it assumes statistical independence of the tests, an assumption that is clearly violated in the context of M/EEG timeseries characterised by massive spatiotemporal dependencies. Some alternative methods aims at controlling the FDR, defined as the proportion of false positive *among positive tests* [e.g., @benjamini1995; @benjamini2001]. However, a major limitation of both types of corrections is that they do not take into account the spatial and temporal information contained in M/EEG data.

A popular technique to account for spatiotemporal dependencies while controlling the FWER is cluster-based inference [@bullmore1999; @maris2007]. A typical cluster-based inference consists of two successive steps [for more details on cluster-based inference, see for instance @frossard2022; @maris2011; @maris2007; @sassenhagen2019]. First, clusters are defined as sets of contiguous timesteps, sensors, voxels, etc, whose activity, summarised by some test statistic (e.g., a $t$-value), exceeds a predefined threshold (e.g., the 95th percentile of the parametric null distribution). Clusters are then characterised by their height (i.e., maximal value), extent (number of constituent elements), or some combination of both, for instance by summing the statistics within a cluster, an approach referred to as "cluster mass" [@maris2007; @pernet2015]. Then, the null hypothesis is tested by computing a $p$-value for each identified cluster by comparing its mass with the null distribution of cluster masses (obtained via permutation). As alluded previously, a significant cluster is a cluster which contains *at least one* significant time-point. As such, it would be incorrect to conclude, for instance, that the timestep of a significant cluster is the first moment at which some conditions differ [@frossard2022; @sassenhagen2019]. Because the inference is performed at the second step (i.e., once clusters have been formed), no conclusion can be made about individual datapoints (e.g., timesteps, sensors, etc).

@@

where $h_{0}$ is typically $0$ and parameters $E$ and $H$ are set a priori (typically to 0.5 and 2, respectively) and control the influence of the extend and height on the TFCE. In practice, this integral is approximated by a sum over small $h$ increments. Then, a *p*-value for each timestep $t$ is computed by comparing its TFCE with the null distribution of TFCE values (obtained via permutation). For each permuted signal, we keep the maximal value over the whole signal for the null distribution of the TFCE. The TFCE combined with permutation (assuming a large enough number of permutations) has been shown to provide accurate FWER control [e.g., @pernet2015]. However, further simulation work showed that cluster-based methods (including TFCE) perform poorly in localising the onset of M/EEG effects [e.g., @rousselet_using_2025; @sassenhagen2019].

@@

Scalp-recorded M/EEG signals capture neural activity originating from various brain regions and are often contaminated by artifacts unrelated to the cognitive processes under investigation. Consequently, analysing M/EEG data necessitates methods that can disentangle task-relevant neural signals from extraneous "noise." A widely adopted technique for this purpose is the estimation of event-related potentials (ERPs), which are stereotyped electrophysiological responses time-locked to specific sensory, cognitive, or motor events. Typically, ERPs are derived by averaging EEG or MEG epochs across multiple trials aligned to the event of interest (e.g., stimulus onset), thereby enhancing the signal-to-noise ratio by attenuating non-time-locked activity. However, this averaging approach has notable limitations: it assumes consistent latency and amplitude across trials and is primarily suited for simple categorical designs. Such assumptions may not hold in more complex experimental paradigms, potentially leading to suboptimal ERP estimations [e.g., @smith2014a].

To overcome the limitations of simple averaging, several model-based approaches for estimating ERPs have been proposed. These methods are motivated by the observation that traditional ERP averaging is mathematically equivalent to fitting an intercept-only linear regression model in a simple categorical design without overlapping events [@smith2014a]. In contrast to simple averaging, regression-based approaches to ERP estimation offer substantially greater flexibility. Notably, they allow for the modelling of both linear and nonlinear effects of continuous predictors, such as word frequency or age [e.g., @smith2014a; @smith2014b; @tremblay2014], and enable the disentangling of overlapping cognitive processes [e.g., @ehinger_unfold_2019; @skukies_brain_2024; @skukies_modelling_2021]. One widely used implementation of this approach is provided by the LIMO EEG toolbox [@pernet2011], which follows a multi-stage analysis pipeline. First, a separate regression model is fit for each datapoint (e.g., each time point and electrode) at the individual participant level to estimate ERP responses. This is followed by group-level statistical analyses of the resulting regression coefficients, often accompanied by corrections for multiple comparisons or cluster-based inference [for recent applied examples, see @dunagan2025; @wüllhorst2025].

Although this framework allows for the inclusion of a wide range of predictors--both continuous and categorical, linear and nonlinear--it still has important limitations. First, fitting separate models for each datapoint ignores the spatiotemporal dependencies inherent in M/EEG data, potentially reducing statistical power and interpretability. Second, the subsequent group-level analyses typically do not account for hierarchical dependencies which could otherwise be addressed through multilevel modelling. Finally, because the output of this procedure is summarised by cluster-based inference, its conclusions remain subject to the limitations discussed in the previous section.

Beyond modelling nonlinear effects of continuous predictors on ERP amplitudes, GAMs have been employed to capture the temporal dynamics of ERPs themselves, effectively modelling the shape of the waveform over time [@abugaber2023; @baayen2018; @meulman2015; @meulman2023]. This approach allows for the estimation of smooth, data-driven functions that characterise how neural responses evolve over time, offering a flexible alternative to traditional linear models. In the following section, we provide a brief introduction to GAMs, highlighting their applicability to M/EEG time series analysis and the advantages they offer over conventional methods.

@@

In generalised additive models, the functional relationship between the predictors and the response variable is decomposed into a sum of low-dimensional non-parametric functions. A typical GAM has the following form:

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this point: *splines* are functions composed of simpler functions. These simpler functions are called *basis functions* (e.g., cubic polynomial, thin-plate) and the set of basis functions is called a *basis*. Each basis function is weighted by its coefficient and the resultant spline is the sum of these weighted basis functions (@fig-intro-gam\ignorespaces A). Splines coefficients are penalised (usually through the square of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function [@miller2025; @wood2017]. In other words, more complex (wiggly) basis functions are automatically penalised.

@@

```{r fig-intro-gam, fig.width = 9, fig.cap = "Different types of GAMs. **A**: GAMs predictions are computed as the weigthed sum (in black) of basis functions (here thin-plate basis functions, in grey). **B**: Constant-effect GAM, with 5 participants in colours and the group-level prediction in black. **C**: Varying-intercept + varying-slope GAMM (with constant smoother). **D**: Varying-intercept + varying-slope + varying-smoother GAMM. In this model, each participant gets its own intercept, slope, and degree of 'wiggliness' (smoother)."}
########################
# simulating some data #
########################

@@

A detailed treatment of the technical underpinnings of GAMs is beyond the scope of this article [see reference books such as @hastie2017; @wood2017]. However, it is worth emphasising that GAMs have been successfully applied to a wide range of time series data across the cognitive sciences, including pupillometry [e.g., @vanrij2019], articulography [e.g., @wieling2018], speech formant dynamics [e.g., @sóskuthy2021], neuroimaging data [e.g., @dinga2021], and event-related potentials [e.g., @abugaber2023; @baayen2018; @meulman2015; @meulman2023]. Their appeal for modelling M/EEG data lies in their ability to flexibly capture the complex shape of ERP waveforms without overfitting, through the use of smooth functions constrained by penalisation. Recent extensions, such as distributional GAMs [@rigby2005; @umlauf2018], allow researchers to model not only the mean structure but also the variance (or scale) and other distributional properties as functions of predictors, a feature that has proven useful in modelling neuroimaging data [e.g., @dinga2021]. Moreover, hierarchical or multilevel GAMs [@pedersen_hierarchical_2019] provide a principled way to account for the nested structure of M/EEG data (e.g., trials within participants), enabling the inclusion of varying intercepts, slopes, and smoothers (as illustrated in @fig-intro-gam\ignorespaces C-D). This approach mitigates the risk of overfitting and reduces the influence of outliers on smooth estimates [@baayen2020; @meulman2023].

@@

Cluster-based permutation tests are widely used in M/EEG research to identify statistically significant effects across time and space. However, these methods have notable limitations, particularly in accurately determining the precise onset and offset of neural effects. To address these limitations, we developed a model-based approach relying on Bayesian generalised additive multilevel models implemented in `R` via the `brms` package [@brms2017; @brms2018]. We evaluated the performance of this approach against conventional methods using both simulated and actual M/EEG data. Our findings demonstrate that this method provides more precise and reliable estimates of effects' onset and offset than conventional approaches such as cluster-based inference.

# Benchmarking with known ground truth

## Methods

### M/EEG data simulation

To assess the accuracy of group-level onset and offset estimation of our proposed method, we simulated EEG with known onset and offset values. Following the approach of @sassenhagen2019 and @rousselet_using_2025, we simulated EEG data stemming from two conditions, one with noise only, and the other with noise + signal. As in previous studies, the noise was generated by superimposing 50 sinusoids at different frequencies, following an EEG-like spectrum [see code in the online supplementary materials and details in @yeung2004]. As in @rousselet_using_2025, the signal was generated from a truncated Gaussian distribution with an objective onset at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated data for a group of 20 participants (with variable true onset) with 50 trials per participant and condition (@fig-eeg). All figures and simulation results can be reproduced using the `R` code available online at: <https://github.com/lnalborczyk/brms_meeg>.

@@

### Model description and model fitting

We then fitted a Bayesian GAM (BGAM) using the `brms` package [@brms2017; @brms2018] and default priors (i.e., weakly informative priors). We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$ [@gabry2019; @gelman2020]. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects. @fig-plot-post-slope shows the predictions of this model together with the raw data.
